---
title: "EDA_ALSPAC"
output:
  pdf_document: default
  html_document: default
date: "2025-06-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(foreign)
library(tidyverse)
```

Reading in ALSPAC data

```{r}
alspac = read.spss("Ulven_B3777_13Jan22.sav",to.data.frame = TRUE,add.undeclared.levels = "no")
#str(alspac)
```

Checking data structure

```{r}
str(alspac)
```

Several variables such as cholesterols are factors and should be integers.

```{r}
alspac = alspac %>%
  mutate(across(c("chol_cord","trig_cord","HDL_cord","LDL_cord","CRP_cord",
                  "Trig_CIF31","Chol_CIF31","LDL_CIF31","HDL_CIF31",
                  "Trig_CIF43","Chol_CIF43","LDL_CIF43","HDL_CIF43",
                  "Chol_BBS","Trig_BBS","LDL_BBS","CHOL_F7","TRIG_F7","HDL_F7",
                  "LDL_F7","CHOL_F9","trig_f9","HDL_f9","LDL_f9","VLDL_f9","CRP_f9","IL6_f9"),as.numeric))
str(alspac)
```


Drop extra factors from variables that are categorical






Count the NA values per sample

```{r}
na_count = apply(alspac,2,function(x){sum(is.na(x))})
na_count_o = sort(na_count,decreasing = TRUE)
na_count_o[1:10]
varibles_half_na = names(na_count_o)[na_count_o >= max(na_count_o)/2]
`%ni%` = Negate(`%in%`)
alspac_use = alspac[,colnames(alspac) %ni% varibles_half_na]
```
Using mice for missing data analysis among remaining variables

```{r}
library(mice)
library(VIM)
## red is missing and blue is not missing
missing_val_plot = aggr(alspac_use[,c(1:10)], col=c("blue","red"),
                   numbers=TRUE,sortVard=TRUE,
                     labels = names(alspac_use),cex.axis=.7,
                       gap=3,ylab=c("porportion of missing","pattern"))



#md.pattern(alspac_use,rotate.names = TRUE)
```
Display missing-data patterns
```{r}
tab.pattern=md.pattern(alspac_use[,c(1:10)],rotate.names = TRUE) ## difficul tot understand
md.pattern(alspac_use[,c(1,11:20)],rotate.names = TRUE)
md.pattern(alspac_use[,c(1,21:26)],rotate.names = TRUE)
```
Doing missing data pattern for the most interesting variables
```{r}
cols_use = c("cidB3777","no2_preg","pm25_preg","pm10_preg","kz021","c800","c804","c645a","b650")
pdf("Missing_pattern_table.pdf",width = 10,height = 10)
md.pattern(alspac_use[,cols_use],rotate.names = TRUE)
dev.off()
```


Both no2_preg and pm25_preg have the most missing values

```{r}
missing_b650<-is.na(alspac_use$b650)#Missing indicator for ever smoked

missing_c645a<-is.na(alspac_use$c645a)#Missing indicator for mom education

missing_c800<-is.na(alspac_use$c800)#Missing indicator for ethnic group

missing_c804<-is.na(alspac_use$c804)#Missing indicator for child ethnicity

missing_areases<-is.na(alspac_use$areases_quint_preg)#Missing indicator for areaSES


```

Using the missing data indicator variable created, let us compare the mean PM2.5 exposure of those with missing values on Length against those whose Length was observed. 

```{r}
tapply(X=alspac_use$pm25_preg,INDEX=missing_b650,FUN=mean,na.rm=T)
tapply(X=alspac_use$pm25_preg,INDEX=missing_c645a,FUN=mean,na.rm=T)
tapply(X=alspac_use$pm25_preg,INDEX=missing_c800,FUN=mean,na.rm=T)
tapply(X=alspac_use$pm25_preg,INDEX=missing_c804,FUN=mean,na.rm=T)
tapply(X=alspac_use$pm25_preg,INDEX=missing_areases,FUN=mean,na.rm=T)


```
Above table suggests no notable difference pm25_preg for those categorical variables.

```{r}
t.test(alspac_use$pm25_preg ~ missing_b650)
t.test(alspac_use$pm25_preg ~ missing_c645a)
t.test(alspac_use$pm25_preg ~ missing_c800)
t.test(alspac_use$pm25_preg ~ missing_c804)
#t.test(alspac_use$pm25_preg ~ missing_areases)

```
There is no significant difference in PM2.5 exposure for those with or without
data on some potential covariates.
Making a jitter plot for PM2.5 by various groups
```{r}
plot(jitter(as.numeric(missing_c804)),alspac_use$pm25_preg,ylim=c(9,16),
            xaxt="n",
            xlab="Mother ever smoked missingness",
            ylab="PM2.5 during pregnancy")
axis(1, at=c(0,1), labels=c("Not missing","Missing"))

abline(h=quantile(alspac_use$pm25_preg,1/4,na.rm=T)-1.5*IQR(alspac_use$pm25_preg,na.rm=T),col="red")
```
Now doing the same for PM10
First checking number of missing Pm10 values by confounders
```{r}
tapply(X=alspac_use$pm10_preg,INDEX=missing_b650,FUN=mean,na.rm=T)
tapply(X=alspac_use$pm10_preg,INDEX=missing_c645a,FUN=mean,na.rm=T)
tapply(X=alspac_use$pm10_preg,INDEX=missing_c800,FUN=mean,na.rm=T)
tapply(X=alspac_use$pm10_preg,INDEX=missing_c804,FUN=mean,na.rm=T)
tapply(X=alspac_use$pm10_preg,INDEX=missing_areases,FUN=mean,na.rm=T)

```

T test for PM10

```{r}

t.test(alspac_use$pm10_preg ~ missing_b650)
t.test(alspac_use$pm10_preg ~ missing_c645a)
t.test(alspac_use$pm10_preg ~ missing_c800)
t.test(alspac_use$pm10_preg ~ missing_c804)
#t.test(alspac_use$pm25_preg ~ missing_areases)

```


```{r}
plot(jitter(as.numeric(missing_b650)),alspac_use$pm10_preg,ylim=c(26,42),
            xaxt="n",
            xlab="Mother ever smoked missingness",
            ylab="PM10 during pregnancy")
axis(1, at=c(0,1), labels=c("Not missing","Missing"))

abline(h=quantile(alspac_use$pm10_preg,1/4,na.rm=T)-1.5*IQR(alspac_use$pm10_preg,na.rm=T),col="red")
```

Now, I will drop the extra levels of all categorical variables that do not have values corresponding to those levels
```{r}
alspac_use_no_extra_factor <- alspac_use %>%
  mutate(across(where(is.factor), fct_drop))
```

Now chekcing if extra factors dropped
```{r}
str(alspac_use)
str(alspac_use_no_extra_factor)
```

Using the missing data indicator variable created, let us compare the mean NO2 exposure of those with missing values on Length against those whose Length was observed. 


```{r}
tapply(X=alspac_use_no_extra_factor$no2_preg,INDEX=missing_b650,FUN=mean,na.rm=T)
tapply(X=alspac_use_no_extra_factor$no2_preg,INDEX=missing_c645a,FUN=mean,na.rm=T)
tapply(X=alspac_use_no_extra_factor$no2_preg,INDEX=missing_c800,FUN=mean,na.rm=T)
tapply(X=alspac_use_no_extra_factor$no2_preg,INDEX=missing_c804,FUN=mean,na.rm=T)
#tapply(X=alspac_use$no2_preg,INDEX=missing_areases,FUN=mean,na.rm=T)

```
T-test for NO2.

```{r}

t.test(alspac_use_no_extra_factor$no2_preg ~ missing_b650)
t.test(alspac_use_no_extra_factor$no2_preg ~ missing_c645a)
t.test(alspac_use_no_extra_factor$no2_preg ~ missing_c800)
t.test(alspac_use_no_extra_factor$no2_preg ~ missing_c804)
#t.test(alspac_use$pm25_preg ~ missing_areases)
```
The above data suggests that NO2 has different distribution by several of the potential covariates.

```{r}
plot(jitter(as.numeric(missing_b650)),alspac_use$no2_preg,ylim=c(15,36),
            xaxt="n",
            xlab="Mother ever smoked missingness",
            ylab="NO2 during pregnancy")
axis(1, at=c(0,1), labels=c("Not missing","Missing"))

abline(h=quantile(alspac_use$no2_preg,1/4,na.rm=T)-1.5*IQR(alspac_use$no2_preg,na.rm=T),col="red")

plot(jitter(as.numeric(missing_c645a)),alspac_use$no2_preg,ylim=c(15,36),
            xaxt="n",
            xlab="Maternal education",
            ylab="NO2 during pregnancy")
axis(1, at=c(0,1), labels=c("Not missing","Missing"))

abline(h=quantile(alspac_use$no2_preg,1/4,na.rm=T)-1.5*IQR(alspac_use$no2_preg,na.rm=T),col="red")

plot(jitter(as.numeric(missing_c800)),alspac_use$no2_preg,ylim=c(15,36),
            xaxt="n",
            xlab="Child ethnic group",
            ylab="NO2 during pregnancy")
axis(1, at=c(0,1), labels=c("Not missing","Missing"))

abline(h=quantile(alspac_use$no2_preg,1/4,na.rm=T)-1.5*IQR(alspac_use$no2_preg,na.rm=T),col="red")

plot(jitter(as.numeric(missing_c804)),alspac_use$no2_preg,ylim=c(15,36),
            xaxt="n",
            xlab="Child ethnicity",
            ylab="NO2 during pregnancy")
axis(1, at=c(0,1), labels=c("Not missing","Missing"))

abline(h=quantile(alspac_use$no2_preg,1/4,na.rm=T)-1.5*IQR(alspac_use$no2_preg,na.rm=T),col="red")

plot(jitter(as.numeric(missing_areases)),alspac_use$no2_preg,ylim=c(15,36),
            xaxt="n",
            xlab="Area SES",
            ylab="NO2 during pregnancy")
axis(1, at=c(0,1), labels=c("Not missing","Missing"))

abline(h=quantile(alspac_use$no2_preg,1/4,na.rm=T)-1.5*IQR(alspac_use$no2_preg,na.rm=T),col="red")


```
Checking if the exposures are different by gender
```{r}
t.test(alspac_use_no_extra_factor$no2_preg ~ alspac_use_no_extra_factor$kz021)
t.test(alspac_use_no_extra_factor$pm25_preg ~ alspac_use_no_extra_factor$kz021)
t.test(alspac_use_no_extra_factor$pm10_preg ~ alspac_use_no_extra_factor$kz021)

```
While distribution of pollutant exposure does not differ by gender, gender
differences lead to differences in expression of genes and hence adjusting for
gender is necessary. This analysis is followed by differential expression analysis
using the bryois.csv microarray matrix from ALSPAC.




